If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-23 GroupPolarizationStatmod/paper/ms.tex
tightening intro

\usepackage{authblk}
\usepackage{etoolbox}
\usepackage{wrapfig}
\makeatletter
\title{If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Environmental Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{\today}
\begin{document}
\maketitle
% \begin{abstract}
% \noindent 
% \emph{Group polarization} is the name for a form of consensus where
% members of a like-minded group become, on average, more extreme in their
% opinions after discussing a topic. Group polarization is important because
% it may increase social and political tensions if moderates become more
% extreme. Decades of studies have replicated detections of group
% polarization, typically using an ordinal scale to measure opinions.
% However, these studies did not account for ordinal measurements, which can
% result in spurious detections of opinion change.  Lacking original data we
% can still calculate the probability that a published detection of group
% polarization was spurious, i.e., to calculate the \emph{false detection
% rate}.  Across 54 group polarization experimental conditions from ten
% representative journal articles, we calculated false detection rates to be
% between 0.52 and 0.88, with a median of 0.75, using a generative model of
% group polarization experiments seeded with empirical data.  
% We also use our model to develop experimental designs that
% achieve an acceptable false discovery rate.  Much of group polarization
% research may be unreliable. This analysis can help change that by enabling others to avoid this
% hidden pitfall. More
% broadly this work demonstrates one important way that replication success alone
% does not imply epistemic reliability.
% \end{abstract}

@@

\begin{quote}
One of the most robust findings in social psychology is that of attitude polarization 
following discussion with like-minded others.
\par\raggedleft(Cooper et al, 2001, p. 267 \cite[p. 267]{Cooper2001})
\end{quote}

\begin{quote}
A medium effect size is conceived as one large enough to be visible to the naked eye.
\par\raggedleft(Cohen, 1988 \cite[p. 26]{CohenBook1988})
\end{quote}

@@

\mt{One more brick to fall in the ongoing replication and generalizability crisis \autocite{Yarkoni2022}.}

@@

Our analysis brings even more additional questions into focus. Most
significantly, recall that even the more appropriate ordered probit Bayesian
model often failed to accurately detect negative results with the number of
participants used in the original papers. Many of the studies analyzed here
did not use any statistical test to establish significance; those that did
used some form of a $t$-test, and it is not clear if the $t$-test that was
used accounted for a change in pre- and post-deliberation opinion variance,
which we expect due to consensus.

The ordered probit model we used still likely underestimates the variance in
outcomes, which could require even larger sample sizes. A more rigorous
analysis would include additional sources of variance that will likely
further inflate the Type I error rate, $\alpha$, and the false discovery
rate~\cite{Yarkoni2022}. First and foremost, there is variance within each
experimental deliberation group; no group polarization study accounts for
this, instead treating participants in each experimental condition as if
they were one large group. Furthermore, survey data are known to be
noisy~\cite{Zaller1992}. People report opinions differently over time for no
apparent reason (i.e., opinions are \emph{unstable}).  Context matters: the
order in which a survey question is asked, and which question framing is
chosen among logically equivalent alternatives, can both significantly
influence participant responses. Finally, little is known about the
psychological process that converts latent opinions to reporting behaviors
(e.g., clicking a radio button corresponding to an opinion). 
The effect of accounting for these sources of variance must be understood to
estimate $\alpha$ (and power, $W$) for the design of the next generation of
group polarization experiments.

To close, note that replication of weakly- and incommensurately-specified 
social phenomena is not enough: strong, mechanistic models must be developed to
be able to predict Type I and II error rates \emph{ahead of time}, as we
demonstrate here. We only demonstrate how to calculate expected Type I error
rates for a measurement design, i.e., the rate of false positives. Low
statistical power exacerbates false detection rates. So, to be confident in
replication, a pre-registered measurement design should use this style of
simulation to estimate the statistical power. It is difficult to incorporate
prior knowledge such as experiment structure without a sufficiently rigorous
and appropriate formal system
\cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.

>>>>

\usepackage{authblk}
\usepackage{etoolbox}
\usepackage{wrapfig}
\makeatletter
% \title{If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization}
\title{If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Environmental Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{\today}
\begin{document}
\maketitle

@@

\begin{quote}
  The good thing about science is that it's true, whether or not you believe
  it. (Neil de Grasse Tyson slogan circa 2018)
\end{quote}

@@

@@

When we used the appropriate ordered probit Bayesian statistical model, it
still often failed to accurately identify mere agreement, instead detecting
group polarization.  The ordered probit model we used still likely
underestimates the variance in outcomes, which could require even larger
sample sizes. A more rigorous analysis would include additional sources of
variance that will likely further inflate the Type I error rate, $\alpha$,
and the false discovery rate~\cite{Yarkoni2022}. First and foremost, there is
variance within each experimental deliberation group; no group polarization
study accounts for this, instead treating participants in each experimental
condition as if they were one large group. Furthermore, survey data are known
to be noisy~\cite{Zaller1992}. People report opinions differently over time
for no apparent reason (i.e., opinions are \emph{unstable}).  Context
matters: the order in which a survey question is asked, and which question
framing is chosen among logically equivalent alternatives, can both
significantly influence participant responses. Finally, little is known about
the psychological process that converts latent opinions to reporting
behaviors (e.g., clicking a radio button corresponding to an opinion).  The
effect of accounting for these sources of variance must be understood to
estimate $\alpha$ (and power, $W$) for the design of the next generation of
group polarization experiments.

A simple change in measurement procedure could fix the problem identified
here: center the post-discussion scale on each group’s own pre-discussion
mean. This removes the assumption that all groups share a common “neutral”
midpoint. The situation is like tossing a ball on a moving train: to the
passengers, the ball’s speed is the same in both directions, but to an
observer on the ground one throw appears faster and the other slower.
Likewise, when measurement is anchored to an external scale, simple agreement
within a group can appear as movement toward one pole. Re-centering the
measurement frame corrects this distortion. It acknowledges that opinion
change is relative to the observer’s reference frame—there is no fixed
background of neutrality against which all groups can be compared. This is,
in essence, the insight of Newtonian relativity: apparent motion depends on
the frame from which it is measured.

First we must note that there are even more, perhaps even graver problems
with group polarization research than I discussed so far: one of note is
opinion instability. Opinions are not generally stable—people seem to
construct their opinions based on several factors not directly related to
whatever underlying topic. Factors affecting reported opinions include the
language used to frame a prompt or question; the order of questions on a
multi-item survey; or simply when the question was asked (Kalton and Schuman,
1982; Zaller and Feldman, 1992; Zaller 1992). Failure to account for sources
of input variance in any experiment can lead to inflated effect sizes since
the input variance is missing from the posterior predictions during
statistical model fitting (Yarkoni, 2022). This further compounds the
theoretical confusions that fill the void when theories are never formalized
and grounded in a mathematical or computational model (Smaldino 2020; Turner
and Smaldino, 2022).

A second additional problem with group polarization research is that we literally cannot learn anything by comparing summary statistics obtained with incommensurate experimental designs—but this is exactly how group polarization research has tried to make progress. Comparing summary statistics only has explanatory power if datasets were generated under the same experimental designs—it seems this was never done or even considered in the group polarization literature. In order to compare studies, their datasets must be generated from commensurate experiments, meaning the studies must share the same predictive and statistical models and experimental design. Otherwise, such comparisons are meaningless, they literally tell us nothing, as shown by philosopher of science Nancy Cartwright in Chapter 5 of her book The Dappled World: A Study of the Boundaries of Science.

\subsection{Humble Hearts Promote Rigorous Science}

Cass Sunstein kicked off the 2000s by elevating group polarization to be a scientific law and advocating the application of group polarization research to serious real-world problems in two books (Sunstein, 2009; Sunstein, 2019) and two recent high-profile articles in Nature Human Behavior: one on the social foundations of pandemic preparedness (Van Bavel, et al., 2020) and one to “promote truth, autonomy, and democratic discourse online” (Lorenz-Spreen, Lewandowski, Sunstein, and Hertwig, 2020). Let us try to stop the spread of this bad science—and let us explore what must be done to achieve Dorian Cartwright’s vision of a rigorous science of group polarization he articulated fifty years ago.

To be practically useful, science must be rigorous. Rigorous science comes most naturally when scientists remain humble, especially when they serve at the pleasure of the public, the government, or other donors. Here’s another example from Burnstein and Vinokur (1973) that we can learn from—let’s try to avoid the sort of arrogance and haughtiness that seep from these acknowledgements they wrote on p. 123:

\begin{quote}
This research was supported by a Grant from the National Institute of Mental Health (MH-16950-03) and by a Guggenheim Foundation Fellowship awarded to the first author. We gratefully acknowledge the incisive, witty, and vinous commentary provided by colleagues at the University of Provence, in particular Robert Abelson, Claude Flament, and Jean-Pierre Poitou, as well as the spirited and intelligent efforts of Livia Mezrick and Irene Graczyk, who helped us carry out the study.
\end{quote}

Science is not glamorous—it takes guts and sacrifice to achieve the technical skills and discipline to sharpen one’s prose, methods, mathematics, and computer code to the point they can really cut reality at its joints, investigate different parts, then put it all back together with a message for others about how the world really works. These days, to do rigorous social science work one must wisely curate the best of the available social science by harmonizing cross-disciplinary jargon. But disincentives still exist against careful scholarship. Prizes and professorships go to those with the most numerous publications in the right journals—period.

Unpaid, anonymous reviewers are supposed to act as gatekeepers to catch problems like the ones I reviewed. But, journal article reviewers have little to gain from enforcing rigor—actually there is a lazy strategy that can pay off big for freeloaders. It’s easy to imagine a cultural norm evolving where reviewers save themselves time and energy by asking little of an author, in hopes that little will be demanded of them when the reviewer submits their next paper. This could result in a publication process that rewards the laziest and most willing to flood the zone and self-aggrandize.

\subsection{Conclusion}

Social science can use physics as a model discipline to transition from idiosyncratic and Balkanized to informative and united. Physics only changed the world because it is useful—it is useful only because it reliably predicts things. Physics became more useful as it became more united, finding shared representations for concepts developed independently—for example the science of optics was developed independently from the science of dynamics, but the physics of light is explained partly in analogy to a pendulum or spring—the mathematics of harmonic motion can be used to predict the motion of springs or pendulums—or why we see different colors of light through a prism. With careful, coordinated tweaks to group polarization experimental design and theoretical modeling, we will build a more rigorous theory of group polarization and its large-scale expression, echo chamber radicalization.

I expect that group polarization experiments, when properly constructed and analyzed, will help sustainability campaigns optimize their social influence campaigns. We can start by developing simpler experimental designs based on formal or computational models capable of predicting outcomes in an experiment. I have developed a model that could motivate an experimental design: I showed that if extremists are more stubborn, this is sufficient to cause group polarization when likeminded groups interact over time. There is even a free stubborn extremism parameter that could be fit to data to measure the relationship between stubbornness and extremism on different issues.

Group polarization research so far has mostly provided us with a checklist of what not to do. The alternative approach I outline builds on rock—the rock of rigor—and will supersede past group polarization research built on a thousand grains of inexact verbal speculation.
